---
title: "Final Project"
author: "Ethan, Stephen, William"
date: "5/6/2021"
output: pdf_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

##Load Packages
library(RSQLite)
library(tidyverse)
library(xgboost)
library(Matrix)
library(knitr)
library(readxl)
```

```{r SQLdatabase}
##Connect

con <- dbConnect(drv=RSQLite::SQLite(), dbname="database.sqlite")

##Getting the tables

tables <- dbListTables(con)
tables <- tables[tables != "sqlite_sequence"]

##Reading in SQL DATA

country = dbReadTable(con, "Country")
league = dbReadTable(con, "League")
matches = dbReadTable(con, "Match")
player = dbReadTable(con, "Player")
player_Attributes = dbReadTable(con, "Player_Attributes")
teams = dbReadTable(con, "Team")
team_attributes = dbReadTable(con, "Team_Attributes")

##Disconnect from Database
dbDisconnect(con)


# Importing dataset made by Stephen 
library(readr)
team_ratings_dataset <- read_csv("dataset.csv")

```

# Data Source

| This data is an SQL database containing information for more than 25,000 matches and 10,000 players from European professional football. 

| The source of the data originates from: 

* http://football-data.mx-api.enetscores.com/ : scores, lineup, team formation and events 

* http://www.football-data.co.uk/ : betting odds.  

* http://sofifa.com/ : players and teams attributes from EA Sports FIFA games. 

This can be found at https://www.kaggle.com/hugomathien/soccer where these are all compiled into one large dataset.

------------------------------------------------------------


\newpage

# Team Rating Database

```{r Data Description 2}
head(team_ratings_dataset)
```

* This is a csv database created by Stephen that calculates the average team rating, derived from the average player ratings of all players on each team. 

* The data used to create these values was all from our orginal dataset. 

* To save computational time, we simply imported the data set instead of running the code every time to create it. This database was created in R, the code used can be found in the file DataCleanseCode.R

\newpage


```{r data viz stuff, echo = FALSE}
### SUBSET RELEVANT INFORMATION FROM MATCHES AND TEAMS
matchesviz <- matches %>%
  select(country_id, home_team_api_id, away_team_api_id, home_team_goal, away_team_goal)
  
teamsviz <- teams %>%
  select(team_api_id, team_long_name)

leagueviz <- league %>%
  select(-id)

### CHANGE TEAM API WITH TEAM NAME
matchesviz <- left_join(matchesviz, teamsviz, by = c("home_team_api_id" = "team_api_id")) %>%
  select(-home_team_api_id) %>%
  rename(home_team = team_long_name) 

matchesviz <- left_join(matchesviz, teamsviz, by = c('away_team_api_id' = "team_api_id")) %>%
  select (-away_team_api_id) %>%
  rename(away_team = team_long_name)

### CHANGE COUNTRY_ID TO LEAGUE TYPE
matchesviz <- left_join(matchesviz, leagueviz) %>%
  select(-country_id) %>%
  rename(league = name)

### CREATING WIN PERCENT VARIABLE
home_win_pct <- matchesviz %>%
  mutate(match_result = ifelse(home_team_goal > away_team_goal, "W", 
                               ifelse(home_team_goal == away_team_goal, "T", "L"))) %>%
  select(home_team, match_result, league) %>%
  group_by(home_team, league) %>%
  summarise(win_pct = sum(match_result == "W")/(sum(match_result == "W") + sum(match_result != "W")))

away_win_pct <- matchesviz %>%
  mutate(match_result = ifelse(away_team_goal > home_team_goal, "W",
                               ifelse(away_team_goal == home_team_goal, "T", "L"))) %>%
  select(away_team, match_result, league) %>%
  group_by(away_team, league) %>%
  summarise(win_pct = sum(match_result == "W")/(sum(match_result == "W") + sum(match_result != "W")))

matches_final <- left_join(home_win_pct, away_win_pct, by = c("home_team" = "away_team")) %>%
  rename(team = home_team,
         league = league.x) %>%
  group_by(team, league) %>%
  summarise(win_pct = (win_pct.x + win_pct.y)/2)

                                    ### BEST PERFORMING TEAMS  FACETED BY LEAGUE

 matches_final %>%
  arrange(desc(win_pct)) %>%
  head(20) %>%
  ggplot(aes(reorder(team, win_pct), win_pct, fill = league)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "Top 20 All-Time Best Teams",
       subtitle = "Aggregated by League",
       x = "Team",
       y = "Historical Win Percentage",
       fill = "League")
                                 

                                    ### WORST PERFORMING TEAMS FACETED BY LEAGUE

matches_final %>%
  arrange(win_pct) %>%
  head(20) %>%
  ggplot(aes(reorder(team, -win_pct), win_pct, fill = league)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "Top 20 All-Time Worst Teams",
       subtitle = "Aggregated by League",
       x = "Team",
       y = "Historical Win Percentage",
       fill = "League")

### CREATING SUMMARY STATS FOR HOME AND AWAY
home_stats <- matchesviz %>%
  group_by(home_team) %>%
  mutate(match_result = ifelse(home_team_goal > away_team_goal, "W",
                               ifelse(home_team_goal == away_team_goal, "T", "L")),
         win_pct = sum(match_result == "W")/(sum(match_result == "W") + sum(match_result != "W")),
         loss_pct = sum(match_result == "L")/(sum(match_result == "L") + sum(match_result != "L")),
         tie_pct = 1 - (win_pct + loss_pct),
         goals_per_game = mean(home_team_goal)) %>%
  select(home_team, league, win_pct, loss_pct, tie_pct, goals_per_game)

away_stats <- matchesviz %>%
  group_by(away_team) %>%
  mutate(match_result = ifelse(away_team_goal > home_team_goal, "W",
                               ifelse(away_team_goal == home_team_goal, "T", "L")),
         win_pct = sum(match_result == "W")/(sum(match_result == "W") + sum(match_result != "W")),
         loss_pct = sum(match_result == "L")/(sum(match_result == "L") + sum(match_result != "L")),
         tie_pct = 1 - (win_pct + loss_pct),
         goals_per_game = mean(away_team_goal)) %>%
  select(away_team, league, win_pct, loss_pct, tie_pct, goals_per_game)

                                      ### SUMMARY STATISTICS

summary_stats <- left_join(home_stats, away_stats, by = c("home_team" = "away_team")) %>%
  rename(team = home_team,
         league = league.x,
         goals_per_game = goals_per_game.x) %>%
  group_by(team, league) %>%
  mutate(win_pct = (win_pct.x + win_pct.y)/2,
         loss_pct = (loss_pct.x + loss_pct.y)/2,
         tie_pct = (tie_pct.x + tie_pct.y)/2) %>%
  select(team, league, win_pct, loss_pct, tie_pct, goals_per_game) %>%
  distinct(team, .keep_all = TRUE)

                                      ### CORRELATION OF HOME WINS AND AWAY WINS (NOW SQUARE)
left_join(home_stats, away_stats, by = c("home_team" = "away_team")) %>%
  rename(team = home_team) %>%
  group_by(team) %>%
  summarise(avg_home_win_pct = mean(win_pct.x),
            avg_away_win_pct = mean(win_pct.y)) %>%
  ggplot(aes(avg_home_win_pct, avg_away_win_pct)) +
  geom_point() +
  coord_fixed (ratio=1)+
  geom_smooth(method = "loess") +
  labs(title = "Correlation Between Home/Away Win Percentages",
       x = "Historical Home Win Percentage",
       y = "Historical Away Win Percentage")



readr::write_csv(summary_stats, file = "summary_stats.csv")



                                  ### Distribution of Win Probabilities
boxplot(summary_stats$win_pct, horizontal=TRUE, main="Distribution of Win Probabilities")

```


``` {r more summary stats, echo = FALSE}

league_avg = team_ratings_dataset %>% group_by(league)  %>% summarise(
  "League_Average" = mean((home_team_rating+away_team_rating)/2)
) 

ggplot(league_avg, aes(league, League_Average,  fill = league)) + 
  geom_bar(stat = "identity") +
   coord_flip() +
   labs(title = "League Average Rating",
        x = "League",
        y = "Average Rating",
        fill = "League"
)



```

# Summary Statstics and Visualizations

| Here shows some visualizations of the data, the first showing the top 20 teams in terms of win percentage and the second showing the bottom 20 teams in terms of win percentage. This should give us a good idea of what to expect out of these teams by the end of the final project. 

| This means that the best team, FC Barcelona, should be predicted to do very well and that the worst team, Cordoba CF, should be predicted to do very poorly. If our model does not reflect this, it will be obvious that we have made an error. 

| In addition to this, we thought it would be interesting to show the correlation between home winning percent and away winning percent. As expected, we see a largely linear line of best fit. This means that a team that wins a lot of home games is likely to also win a lot of away games. However, there were some irregularities at low winning percentage caused by sparse observations and outliers.

| Next, we look at the average rating of each league. From the graph, we can see that Spain LIGA BBVA and England Premier League have the highest average ratings, whereas Switzerland Super League and Poland Ekstraklasa have the lowest. 

| Lastly, we wanted to look at the distribution of win probabilities as you recommended in the Data Analysis. As you can see, there are some very extreme outliers with a very high win percentages, which is something that we really did not expect.

------------------------------------------------------------

\newpage


```{r DataCleaning, include = FALSE}



                                   ###MATCHES DATA



matchesclean = select(matches, "id", "country_id", "league_id", "season", "date", 
"match_api_id", "home_team_api_id", "away_team_api_id", "home_team_goal", "away_team_goal", "goal", "shoton", "shotoff", "foulcommit", "card", "cross", "corner", "possession")

head(matchesclean)

## Getting the team names in 


matches1 <- left_join(matchesclean, teams, by = c("home_team_api_id" = "team_api_id"))

colnames(matches1)[colnames(matches1) == "id.x"] = "match_id"
colnames(matches1)[colnames(matches1) == "id.y"] = "home_team_id"

head(matches1)

matches1 <- left_join(matches1, teams, by = c("away_team_api_id" = "team_api_id"))

head(matches1)

colnames(matches1)[colnames(matches1) == "id"] = "away_team_id"
colnames(matches1)[colnames(matches1) == "team_long_name.y"] = "away_team_name"
colnames(matches1)[colnames(matches1) == "team_short_name.y"] = "away_team_name_S"

colnames(matches1)[colnames(matches1) == "team_long_name.x"] = "home_team_name"
colnames(matches1)[colnames(matches1) == "team_short_name.x"] = "home_team_name_S"

head(matches1)

## Finding the match score

  ##Creating the match score column

    matches_wtname = matches1
    matches_wtname$match_score = NA
    
    head(matches_wtname)
    
    matches_wresult = transform(matches_wtname, match_score = ifelse(home_team_goal > away_team_goal, "Win", ifelse(home_team_goal == away_team_goal, "Tie", "Loss")))
    
    head(matches_wresult)


    
matchesclean1 = select(matches_wresult, "match_id", "country_id", "league_id", "season", "date", "match_api_id", "home_team_api_id", "away_team_api_id", "home_team_goal", "away_team_goal", "match_score", "home_team_name", "away_team_name", "home_team_name_S", "away_team_name_S")

head(matchesclean1)

##Getting country to main dataset 

SData <- left_join(matchesclean1, country, by = c("country_id" = "id"))

SData = select(SData, -country_id)
colnames(SData)[colnames(SData) == "name"] = "country"

head(SData)
    


                             ###TEAM ATTRIBUTES DATA

head(team_attributes)
                    

#switch data to season


  ##create season column

team_attributes$season = NA

head(team_attributes)

teama = transform(team_attrbutes, season = 
                              ifelse(date == '2010-02-22 00:00:00', "2009/2010", 
                              ifelse(date == '2011-02-22 00:00:00', "2010/2011", 
                              ifelse(date == '2012-02-22 00:00:00', "2011/2012",
                              ifelse(date == '2013-09-20 00:00:00', "2013/2014", 
                              ifelse(date == '2014-09-20 00:00:00', "2014/2015", 
                              ifelse(date == '2014-09-19 00:00:00', "2014/2015", 
                              ifelse(date == '2015-09-10 00:00:00', "2015/2016",
                              ifelse(date == '2015-09-20 00:00:00', "2015/2016", NA
)))))))))

head(teama)
teamaHOME = teama
teamaAWAY = teama
colnames(teamaHOME)[colnames(teamaHOME) == "team_api_id"] = "home_team_api_id"


Dataset <- left_join(teamaHOME, SData, by = c('home_team_api_id', 'season'))

colnames(teamaAWAY)[colnames(teamaAWAY) == "team_api_id"] = "away_team_api_id"

Dataset <- left_join(teamaAWAY, SData, by = c('away_team_api_id', 'season'))

head(Dataset)
```



# Data Description 

```{r Data Description}
Data_Description<-read_excel('Data Description.xlsx')
print(Data_Description, n=Inf)
```


* This is a xlsx that William created in excel that had a description of all of the variables that was made from scratch. 

* For clarity we also attached the xlsx document to our submission

------------------------------------------------------------

In order to help with some of the code for data cleaning, we referenced:

* https://www.kaggle.com/abharg16/predicting-epl-team-season-win-percentages/data


```{r preprocessing}
                    ### COMBINE DATASETS ###
Dataset <- Dataset %>%
  left_join(team_ratings %>% select(match_api_id, home_team_rating, away_team_rating),
            by = "match_api_id")

Dataset <- bind_cols(Dataset %>% arrange(match_api_id),
                    team_win_streaks %>% arrange(match_api_id)) %>%
  rename(match_api_id = match_api_id...30) %>%
  select(-match_api_id...42)

Dataset <- Dataset %>%
  relocate(match_score)

                    ### FEATURE ENGINEERING FOR THINGS THAT WONT WORK IN RECIPE
## DROP MISSING MATCHES
Dataset <- Dataset %>%
  drop_na()

## CONVERT CHARACTERS TO FACTORS
Dataset <- Dataset %>%
  mutate_if(is.character, as.factor) %>%
  mutate(match_score = as.factor(match_score))

## REMOVE UNWANTED FEATURES
Dataset <- Dataset %>%
  select(-id, -team_fifa_api_id, -away_team_api_id, -date.x, -season, -match_id, -league_id, -date.y,
         -match_api_id, -home_team_api_id, -home_team_name, -away_team_name, -home_team_api_id,
         -home_team_name_S, -away_team_name_S, -country, -home_team_goal, -away_team_goal,
         -buildUpPlayDribbling)

```

## XGBoost Model

```{r xgboost_model}
                    ### BUILDING MODEL ###
## SPLITTING DATA
set.seed(1)
soccer_split <- initial_split(Dataset, strata = match_score)
soccer_train <- training(soccer_split)
soccer_test <- testing(soccer_split)

## MODEL SPECIFICATIONS
xgb_model <- boost_tree(
  trees = 1000, 
  tree_depth = tune(), min_n = tune(), 
  loss_reduction = tune(),                     
  sample_size = tune(), mtry = tune(),         
  learn_rate = tune(),                         
) %>% 
  set_engine("xgboost") %>% 
  set_mode("classification")

xgb_model

## GRID SPECIFICATIONS
xgb_grid <- grid_latin_hypercube(
  tree_depth(),
  min_n(),
  loss_reduction(),
  sample_size = sample_prop(),
  finalize(mtry(), soccer_train),
  learn_rate(),
  size = 30
)

kable(head(xgb_grid))

## SETTING UP WORKFLOW
xgb_workflow <- workflow() %>%
  add_formula(match_score ~ .) %>%
  add_model(xgb_model)

xgb_workflow

## FOLDS
set.seed(1)
vb_folds <- vfold_cv(soccer_train, strata = match_score)

vb_folds

## PRAY FOR MY PC
doParallel::registerDoParallel()

## TUNE GRID
set.seed(1)
xgb_results <- tune_grid(
  xgb_workflow,
  resamples = vb_folds,
  grid = xgb_grid,
  control = control_grid(save_pred = TRUE)
)

view(xgb_results$.notes)

xgb_results

xgb_results %>% collect_metrics()

## VISUALIZING ACCURACY METRICS
xgb_results %>%
  collect_metrics() %>%
  filter(.metric == "roc_auc") %>%
  select(mean, mtry:sample_size) %>%
  pivot_longer(mtry:sample_size,
               values_to = "value",
               names_to = "parameter"
  ) %>%
  ggplot(aes(value, mean, color = parameter)) +
  geom_point(alpha = 0.8, show.legend = FALSE) +
  facet_wrap(~parameter, scales = "free_x") +
  labs(x = NULL, y = "AUC")

## BEST PERFORMING SET OF PARAMETERS
show_best(xgb_results, "accuracy")

## SELECT BEST ACCURACY MEASUREMENT
best_accuracy <- select_best(xgb_results, "accuracy")
best_accuracy

## FINALIZE WORKFLOW
xgb_final <- finalize_workflow(
  xgb_workflow,
  best_accuracy
)

## VIP PLOT
xgb_final %>%
  fit(data = soccer_train) %>%
  pull_workflow_fit() %>%
  vip(geom = "point")

## FIT TO TRAINING DATA
final_results <- last_fit(xgb_final, soccer_split)
collect_metrics(final_results)

#### STUFF TO SAVE
saveRDS(xgb_results, "./xgb_results.rds")
saveRDS(xgb_final, "./xgb_final.rds")
```

#Final Summary of Project
In the end, we got a 76.8% accuracy, which for the scope of the project certainly isn't too bad. We believe the model was slightly overfit, but given the amount of time that this project has taken (much more than we expected), we chose to continue with this model. If we had more time, we would work on trying to make this model even better and maybe even do some more fine tuning additional parameters. 
