---
title: "Cleaned up ExploratoryDA"
author: "Ethan, Stephen, William"
date: "4/19/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

##Load Packages
library(RSQLite)
library(tidyverse)
library(xgboost)
library(Matrix)
library(knitr)
library(readxl)
```

```{r SQLdatabase}
##Connect

con <- dbConnect(drv=RSQLite::SQLite(), dbname="database.sqlite")

##Getting the tables

tables <- dbListTables(con)
tables <- tables[tables != "sqlite_sequence"]

##Reading in SQL DATA

country = dbReadTable(con, "Country")
league = dbReadTable(con, "League")
matches = dbReadTable(con, "Match")
player = dbReadTable(con, "Player")
player_Attributes = dbReadTable(con, "Player_Attributes")
teams = dbReadTable(con, "Team")
team_attributes = dbReadTable(con, "Team_Attributes")

##Disconnect from Database
dbDisconnect(con)


```

# Data Source

| This data is an SQL database containing information for more than 25,000 matches and 10,000 players from European professional football. 

| The source of the data originates from: 

* http://football-data.mx-api.enetscores.com/ : scores, lineup, team formation and events 

* http://www.football-data.co.uk/ : betting odds.  

* http://sofifa.com/ : players and teams attributes from EA Sports FIFA games. 

This can be found at https://www.kaggle.com/hugomathien/soccer where these are all compiled into one large dataset.

------------------------------------------------------------


\newpage


```{r data viz stuff, echo = FALSE}
### SUBSET RELEVANT INFORMATION FROM MATCHES AND TEAMS
matchesviz <- matches %>%
  select(country_id, home_team_api_id, away_team_api_id, home_team_goal, away_team_goal)
  
teamsviz <- teams %>%
  select(team_api_id, team_long_name)

leagueviz <- league %>%
  select(-id)

### CHANGE TEAM API WITH TEAM NAME
matchesviz <- left_join(matchesviz, teamsviz, by = c("home_team_api_id" = "team_api_id")) %>%
  select(-home_team_api_id) %>%
  rename(home_team = team_long_name) 

matchesviz <- left_join(matchesviz, teamsviz, by = c('away_team_api_id' = "team_api_id")) %>%
  select (-away_team_api_id) %>%
  rename(away_team = team_long_name)

### CHANGE COUNTRY_ID TO LEAGUE TYPE
matchesviz <- left_join(matchesviz, leagueviz) %>%
  select(-country_id) %>%
  rename(league = name)

### CREATING WIN PERCENT VARIABLE
home_win_pct <- matchesviz %>%
  mutate(match_result = ifelse(home_team_goal > away_team_goal, "W", 
                               ifelse(home_team_goal == away_team_goal, "T", "L"))) %>%
  select(home_team, match_result, league) %>%
  group_by(home_team, league) %>%
  summarise(win_pct = sum(match_result == "W")/(sum(match_result == "W") + sum(match_result != "W")))

away_win_pct <- matchesviz %>%
  mutate(match_result = ifelse(away_team_goal > home_team_goal, "W",
                               ifelse(away_team_goal == home_team_goal, "T", "L"))) %>%
  select(away_team, match_result, league) %>%
  group_by(away_team, league) %>%
  summarise(win_pct = sum(match_result == "W")/(sum(match_result == "W") + sum(match_result != "W")))

matches_final <- left_join(home_win_pct, away_win_pct, by = c("home_team" = "away_team")) %>%
  rename(team = home_team,
         league = league.x) %>%
  group_by(team, league) %>%
  summarise(win_pct = (win_pct.x + win_pct.y)/2)

                                    ### BEST PERFORMING TEAMS

 matches_final %>%
  arrange(desc(win_pct)) %>%
  head(20) %>%
  ggplot(aes(reorder(team, win_pct), win_pct, fill = league)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "Top 20 All-Time Best Teams",
       subtitle = "Aggregated by League",
       x = "Team",
       y = "Historical Win Percentage",
       fill = "League")

                                    ### WORST PERFORMING TEAMS

matches_final %>%
  arrange(win_pct) %>%
  head(20) %>%
  ggplot(aes(reorder(team, -win_pct), win_pct, fill = league)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "Top 20 All-Time Worst Teams",
       subtitle = "Aggregated by League",
       x = "Team",
       y = "Historical Win Percentage",
       fill = "League")

### CREATING SUMMARY STATS FOR HOME AND AWAY
home_stats <- matchesviz %>%
  group_by(home_team) %>%
  mutate(match_result = ifelse(home_team_goal > away_team_goal, "W",
                               ifelse(home_team_goal == away_team_goal, "T", "L")),
         win_pct = sum(match_result == "W")/(sum(match_result == "W") + sum(match_result != "W")),
         loss_pct = sum(match_result == "L")/(sum(match_result == "L") + sum(match_result != "L")),
         tie_pct = 1 - (win_pct + loss_pct),
         goals_per_game = mean(home_team_goal)) %>%
  select(home_team, league, win_pct, loss_pct, tie_pct, goals_per_game)

away_stats <- matchesviz %>%
  group_by(away_team) %>%
  mutate(match_result = ifelse(away_team_goal > home_team_goal, "W",
                               ifelse(away_team_goal == home_team_goal, "T", "L")),
         win_pct = sum(match_result == "W")/(sum(match_result == "W") + sum(match_result != "W")),
         loss_pct = sum(match_result == "L")/(sum(match_result == "L") + sum(match_result != "L")),
         tie_pct = 1 - (win_pct + loss_pct),
         goals_per_game = mean(away_team_goal)) %>%
  select(away_team, league, win_pct, loss_pct, tie_pct, goals_per_game)

                                      ### SUMMARY STATISTICS

summary_stats <- left_join(home_stats, away_stats, by = c("home_team" = "away_team")) %>%
  rename(team = home_team,
         league = league.x,
         goals_per_game = goals_per_game.x) %>%
  group_by(team, league) %>%
  mutate(win_pct = (win_pct.x + win_pct.y)/2,
         loss_pct = (loss_pct.x + loss_pct.y)/2,
         tie_pct = (tie_pct.x + tie_pct.y)/2) %>%
  select(team, league, win_pct, loss_pct, tie_pct, goals_per_game) %>%
  distinct(team, .keep_all = TRUE)

                                        ### CORRELATION OF HOME WINS AND AWAY WINS
left_join(home_stats, away_stats, by = c("home_team" = "away_team")) %>%
  rename(team = home_team) %>%
  group_by(team) %>%
  summarise(avg_home_win_pct = mean(win_pct.x),
         avg_away_win_pct = mean(win_pct.y)) %>%
  ggplot(aes(avg_home_win_pct, avg_away_win_pct)) +
  geom_point() +
  geom_smooth(method = "loess") +
  labs(title = "Correlation Between Home/Away Win Percentages",
       x = "Historical Home Win Percentage",
       y = "Historical Away Win Percentage")

readr::write_csv(summary_stats, file = "summary_stats.csv")
```

# Summary Statstics and Visualizations

| Here shows some visualizations of the data, the first showing the top 20 teams in terms of win percentage and the second showing the bottom 20 teams in terms of win percentage. This should give us a good idea of what to expect out of these teams by the end of the final project. 

| This means that the best team, FC Barcelona, should be predicted to do very well and that the worst team, Cordoba CF, should be predicted to do very poorly. If our model does not reflect this, it will be obvious that we have made an error. 

| In addition to this, we thought it would be interesting to show the correlation between home winning percent and away winning percent. As expected, we see a largely linear line of best fit. This means that a team that wins a lot of home games is likely to also win a lot of away games. However, there were some irregularities at low winning percentage caused by sparse observations and outliers.

------------------------------------------------------------

\newpage


```{r DataCleaning, include = FALSE}



                                   ###MATCHES DATA



matchesclean = select(matches, "id", "country_id", "league_id", "season", "date", 
"match_api_id", "home_team_api_id", "away_team_api_id", "home_team_goal", "away_team_goal", "goal", "shoton", "shotoff", "foulcommit", "card", "cross", "corner", "possession")

head(matchesclean)

## Getting the team names in 


matches1 <- left_join(matchesclean, teams, by = c("home_team_api_id" = "team_api_id"))

colnames(matches1)[colnames(matches1) == "id.x"] = "match_id"
colnames(matches1)[colnames(matches1) == "id.y"] = "home_team_id"

head(matches1)

matches1 <- left_join(matches1, teams, by = c("away_team_api_id" = "team_api_id"))

head(matches1)

colnames(matches1)[colnames(matches1) == "id"] = "away_team_id"
colnames(matches1)[colnames(matches1) == "team_long_name.y"] = "away_team_name"
colnames(matches1)[colnames(matches1) == "team_short_name.y"] = "away_team_name_S"

colnames(matches1)[colnames(matches1) == "team_long_name.x"] = "home_team_name"
colnames(matches1)[colnames(matches1) == "team_short_name.x"] = "home_team_name_S"

head(matches1)

## Finding the match score

  ##Creating the match score column

    matches_wtname = matches1
    matches_wtname$match_score = NA
    
    head(matches_wtname)
    
    matches_wresult = transform(matches_wtname, match_score = ifelse(home_team_goal > away_team_goal, "Win", ifelse(home_team_goal == away_team_goal, "Tie", "Loss")))
    
    head(matches_wresult)


    
matchesclean1 = select(matches_wresult, "match_id", "country_id", "league_id", "season", "date", "match_api_id", "home_team_api_id", "away_team_api_id", "home_team_goal", "away_team_goal", "match_score", "home_team_name", "away_team_name", "home_team_name_S", "away_team_name_S")

head(matchesclean1)

##Getting country to main dataset 

SData <- left_join(matchesclean1, country, by = c("country_id" = "id"))

SData = select(SData, -country_id)
colnames(SData)[colnames(SData) == "name"] = "country"

head(SData)
    


                             ###TEAM ATTRIBUTES DATA

head(team_attributes)
                    

#switch data to season


  ##create season column

team_attributes$season = NA

head(team_attributes)

teama = transform(team_attributes, season = 
                              ifelse(date == '2010-02-22 00:00:00', "2009/2010", 
                              ifelse(date == '2011-02-22 00:00:00', "2010/2011", 
                              ifelse(date == '2012-02-22 00:00:00', "2011/2012",
                              ifelse(date == '2013-09-20 00:00:00', "2013/2014", 
                              ifelse(date == '2014-09-20 00:00:00', "2014/2015", 
                              ifelse(date == '2014-09-19 00:00:00', "2014/2015", 
                              ifelse(date == '2015-09-10 00:00:00', "2015/2016",
                              ifelse(date == '2015-09-20 00:00:00', "2015/2016", NA
)))))))))

head(teama)
teamaHOME = teama
teamaAWAY = teama
colnames(teamaHOME)[colnames(teamaHOME) == "team_api_id"] = "home_team_api_id"


Dataset <- left_join(teamaHOME, SData, by = c('home_team_api_id', 'season'))

colnames(teamaAWAY)[colnames(teamaAWAY) == "team_api_id"] = "away_team_api_id"

Dataset <- left_join(teamaAWAY, SData, by = c('away_team_api_id', 'season'))

head(Dataset)
```



# Data Description 

```{r Data Description}
Data_Description<-read_excel('Data Description.xlsx')
print(Data_Description, n=Inf)
```


* This was a xlsx that we created in excel that had a description of all of the variables that was made from scratch. 

* For clarity we also attached the xlsx document to our submission

------------------------------------------------------------

In order to help with some of the code for data cleaning, we referenced:

* https://www.kaggle.com/abharg16/predicting-epl-team-season-win-percentages/data



