---
title: "Soccer Project Updated"
author: "Ethan Gerak"
date: "4/16/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

##Load Packages
library(RSQLite)
library(tidyverse)
library(tidymodels)
library(DataExplorer)
library(xgboost)
library(Matrix)
library(knitr)
library(vip)
```

```{r team_rating}
team_ratings <- read_csv("~/SoccerData/team_ratings.csv")
team_win_streaks <- read_csv("~/SoccerData/team_win_streaks.csv")
```

```{r SQLdatabase}
##Connect

con <- dbConnect(drv=RSQLite::SQLite(), dbname="database.sqlite")

##Getting the tables

tables <- dbListTables(con)
tables <- tables[tables != "sqlite_sequence"]

##Reading in SQL DATA

country = dbReadTable(con, "Country")
league = dbReadTable(con, "League")
matches = dbReadTable(con, "Match")
player = dbReadTable(con, "Player")
player_Attributes = dbReadTable(con, "Player_Attributes")
teams = dbReadTable(con, "Team")
team_attributes = dbReadTable(con, "Team_Attributes")

##Disconnect from Database
dbDisconnect(con)


```

```{r data viz stuff}
### SUBSET RELEVANT INFORMATION FROM MATCHES AND TEAMS
matchesviz <- matches %>%
  select(country_id, home_team_api_id, away_team_api_id, home_team_goal, away_team_goal)
  
teamsviz <- teams %>%
  select(team_api_id, team_long_name)

leagueviz <- league %>%
  select(-id)

### CHANGE TEAM API WITH TEAM NAME
matchesviz <- left_join(matchesviz, teamsviz, by = c("home_team_api_id" = "team_api_id")) %>%
  select(-home_team_api_id) %>%
  rename(home_team = team_long_name) 

matchesviz <- left_join(matchesviz, teamsviz, by = c('away_team_api_id' = "team_api_id")) %>%
  select (-away_team_api_id) %>%
  rename(away_team = team_long_name)

### CHANGE COUNTRY_ID TO LEAGUE TYPE
matchesviz <- left_join(matchesviz, leagueviz) %>%
  select(-country_id) %>%
  rename(league = name)

### CREATING WIN PERCENT VARIABLE
home_win_pct <- matchesviz %>%
  mutate(match_result = ifelse(home_team_goal > away_team_goal, "W", 
                               ifelse(home_team_goal == away_team_goal, "T", "L"))) %>%
  select(home_team, match_result, league) %>%
  group_by(home_team, league) %>%
  summarise(win_pct = sum(match_result == "W")/(sum(match_result == "W") + sum(match_result != "W")))

away_win_pct <- matchesviz %>%
  mutate(match_result = ifelse(away_team_goal > home_team_goal, "W",
                               ifelse(away_team_goal == home_team_goal, "T", "L"))) %>%
  select(away_team, match_result, league) %>%
  group_by(away_team, league) %>%
  summarise(win_pct = sum(match_result == "W")/(sum(match_result == "W") + sum(match_result != "W")))

matches_final <- left_join(home_win_pct, away_win_pct, by = c("home_team" = "away_team")) %>%
  rename(team = home_team,
         league = league.x) %>%
  group_by(team, league) %>%
  summarise(win_pct = (win_pct.x + win_pct.y)/2)

                                    ### BEST PERFORMING TEAMS

 matches_final %>%
  arrange(desc(win_pct)) %>%
  head(20) %>%
  ggplot(aes(reorder(team, win_pct), win_pct, fill = league)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "Top 20 All-Time Best Teams",
       subtitle = "Aggregated by League",
       x = "Team",
       y = "Historical Win Percentage",
       fill = "League")

                                    ### WORST PERFORMING TEAMS

matches_final %>%
  arrange(win_pct) %>%
  head(20) %>%
  ggplot(aes(reorder(team, -win_pct), win_pct, fill = league)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "Top 20 All-Time Worst Teams",
       subtitle = "Aggregated by League",
       x = "Team",
       y = "Historical Win Percentage",
       fill = "League")

### CREATING SUMMARY STATS FOR HOME AND AWAY
home_stats <- matchesviz %>%
  group_by(home_team) %>%
  mutate(match_result = ifelse(home_team_goal > away_team_goal, "W",
                               ifelse(home_team_goal == away_team_goal, "T", "L")),
         win_pct = sum(match_result == "W")/(sum(match_result == "W") + sum(match_result != "W")),
         loss_pct = sum(match_result == "L")/(sum(match_result == "L") + sum(match_result != "L")),
         tie_pct = 1 - (win_pct + loss_pct),
         goals_per_game = mean(home_team_goal)) %>%
  select(home_team, league, win_pct, loss_pct, tie_pct, goals_per_game)

away_stats <- matchesviz %>%
  group_by(away_team) %>%
  mutate(match_result = ifelse(away_team_goal > home_team_goal, "W",
                               ifelse(away_team_goal == home_team_goal, "T", "L")),
         win_pct = sum(match_result == "W")/(sum(match_result == "W") + sum(match_result != "W")),
         loss_pct = sum(match_result == "L")/(sum(match_result == "L") + sum(match_result != "L")),
         tie_pct = 1 - (win_pct + loss_pct),
         goals_per_game = mean(away_team_goal)) %>%
  select(away_team, league, win_pct, loss_pct, tie_pct, goals_per_game)

                                      ### SUMMARY STATISTICS

summary_stats <- left_join(home_stats, away_stats, by = c("home_team" = "away_team")) %>%
  rename(team = home_team,
         league = league.x,
         goals_per_game = goals_per_game.x) %>%
  group_by(team, league) %>%
  mutate(win_pct = (win_pct.x + win_pct.y)/2,
         loss_pct = (loss_pct.x + loss_pct.y)/2,
         tie_pct = (tie_pct.x + tie_pct.y)/2) %>%
  select(team, league, win_pct, loss_pct, tie_pct, goals_per_game) %>%
  distinct(team, .keep_all = TRUE)

                                        ### CORRELATION OF HOME WINS AND AWAY WINS
left_join(home_stats, away_stats, by = c("home_team" = "away_team")) %>%
  rename(team = home_team) %>%
  group_by(team) %>%
  summarise(avg_home_win_pct = mean(win_pct.x),
         avg_away_win_pct = mean(win_pct.y)) %>%
  ggplot(aes(avg_home_win_pct, avg_away_win_pct)) +
  geom_point() +
  geom_smooth(method = "loess") +
  labs(title = "Correlation Between Home/Away Win Percentages",
       x = "Historical Home Win Percentage",
       y = "Historical Away Win Percentage")

readr::write_csv(summary_stats, file = "summary_stats.csv")
```

```{r DataCleaning}



                                   ###MATCHES DATA



matchesclean = select(matches, "id", "country_id", "league_id", "season", "date", 
"match_api_id", "home_team_api_id", "away_team_api_id", "home_team_goal", "away_team_goal", "goal", "shoton", "shotoff", "foulcommit", "card", "cross", "corner", "possession")

head(matchesclean)

## Getting the team names in 


matches1 <- left_join(matchesclean, teams, by = c("home_team_api_id" = "team_api_id"))

colnames(matches1)[colnames(matches1) == "id.x"] = "match_id"
colnames(matches1)[colnames(matches1) == "id.y"] = "home_team_id"

head(matches1)

matches1 <- left_join(matches1, teams, by = c("away_team_api_id" = "team_api_id"))

head(matches1)

colnames(matches1)[colnames(matches1) == "id"] = "away_team_id"
colnames(matches1)[colnames(matches1) == "team_long_name.y"] = "away_team_name"
colnames(matches1)[colnames(matches1) == "team_short_name.y"] = "away_team_name_S"

colnames(matches1)[colnames(matches1) == "team_long_name.x"] = "home_team_name"
colnames(matches1)[colnames(matches1) == "team_short_name.x"] = "home_team_name_S"

head(matches1)

## Finding the match score

  ##Creating the match score column

    matches_wtname = matches1
    matches_wtname$match_score = NA
    
    head(matches_wtname)
    
    matches_wresult = transform(matches_wtname, match_score = ifelse(home_team_goal > away_team_goal, "Win", ifelse(home_team_goal == away_team_goal, "Tie", "Loss")))
    
    head(matches_wresult)


    
matchesclean1 = select(matches_wresult, "match_id", "country_id", "league_id", "season", "date", "match_api_id", "home_team_api_id", "away_team_api_id", "home_team_goal", "away_team_goal", "match_score", "home_team_name", "away_team_name", "home_team_name_S", "away_team_name_S")

head(matchesclean1)

##Getting country to main dataset 

SData <- left_join(matchesclean1, country, by = c("country_id" = "id"))

SData = select(SData, -country_id)
colnames(SData)[colnames(SData) == "name"] = "country"

head(SData)
    


                             ###TEAM ATTRIBUTES DATA

head(team_attributes)
                    

#switch data to season


  ##create season column

team_attributes$season = NA

head(team_attributes)

teama = transform(team_attributes, season = 
                              ifelse(date == '2010-02-22 00:00:00', "2009/2010", 
                              ifelse(date == '2011-02-22 00:00:00', "2010/2011", 
                              ifelse(date == '2012-02-22 00:00:00', "2011/2012",
                              ifelse(date == '2013-09-20 00:00:00', "2013/2014", 
                              ifelse(date == '2014-09-20 00:00:00', "2014/2015", 
                              ifelse(date == '2014-09-19 00:00:00', "2014/2015", 
                              ifelse(date == '2015-09-10 00:00:00', "2015/2016",
                              ifelse(date == '2015-09-20 00:00:00', "2015/2016", NA
)))))))))

head(teama)
select(teama, -date)
teamaHOME = teama
teamaAWAY = teama
colnames(teamaHOME)[colnames(teamaHOME) == "team_api_id"] = "home_team_api_id"


Dataset <- left_join(teamaHOME, SData, by = c('home_team_api_id', 'season'))

colnames(teamaAWAY)[colnames(teamaAWAY) == "team_api_id"] = "away_team_api_id"

Dataset <- left_join(teamaAWAY, SData, by = c('away_team_api_id', 'season'))

head(Dataset)
```

#https://www.kaggle.com/abharg16/predicting-epl-team-season-win-percentages/data




##Changes the id to "match_ID"
colnames(matches)[colnames(matches)=="id"] <- "match_ID"

```{r preprocessing}
                    ### COMBINE DATASETS ###
Dataset <- Dataset %>%
  left_join(team_ratings %>% select(match_api_id, home_team_rating, away_team_rating),
            by = "match_api_id")

Dataset <- bind_cols(Dataset %>% arrange(match_api_id),
                    team_win_streaks %>% arrange(match_api_id)) %>%
  rename(match_api_id = match_api_id...30) %>%
  select(-match_api_id...42)

Dataset <- Dataset %>%
  relocate(match_score)

                    ### FEATURE ENGINEERING FOR THINGS THAT WONT WORK IN RECIPE
## DROP MISSING MATCHES
Dataset <- Dataset %>%
  drop_na()

## CONVERT CHARACTERS TO FACTORS
Dataset <- Dataset %>%
  mutate_if(is.character, as.factor) %>%
  mutate(match_score = as.factor(match_score))

## REMOVE UNWANTED FEATURES
Dataset <- Dataset %>%
  select(-id, -team_fifa_api_id, -away_team_api_id, -date.x, -season, -match_id, -league_id, -date.y,
         -match_api_id, -home_team_api_id, -home_team_name, -away_team_name, -home_team_api_id,
         -home_team_name_S, -away_team_name_S, -country, -home_team_goal, -away_team_goal,
         -buildUpPlayDribbling)

```

## XGBoost Model

```{r xgboost_model}
                    ### BUILDING MODEL ###
## SPLITTING DATA
set.seed(1)
soccer_split <- initial_split(Dataset, strata = match_score)
soccer_train <- training(soccer_split)
soccer_test <- testing(soccer_split)

## MODEL SPECIFICATIONS
xgb_model <- boost_tree(
  trees = 1000, 
  tree_depth = tune(), min_n = tune(), 
  loss_reduction = tune(),                     
  sample_size = tune(), mtry = tune(),         
  learn_rate = tune(),                         
) %>% 
  set_engine("xgboost") %>% 
  set_mode("classification")

xgb_model

## GRID SPECIFICATIONS
xgb_grid <- grid_latin_hypercube(
  tree_depth(),
  min_n(),
  loss_reduction(),
  sample_size = sample_prop(),
  finalize(mtry(), soccer_train),
  learn_rate(),
  size = 30
)

kable(head(xgb_grid))

## SETTING UP WORKFLOW
xgb_workflow <- workflow() %>%
  add_formula(match_score ~ .) %>%
  add_model(xgb_model)

xgb_workflow

## FOLDS
set.seed(1)
vb_folds <- vfold_cv(soccer_train, strata = match_score)

vb_folds

## PRAY FOR MY PC
doParallel::registerDoParallel()

## TUNE GRID
set.seed(1)
xgb_results <- tune_grid(
  xgb_workflow,
  resamples = vb_folds,
  grid = xgb_grid,
  control = control_grid(save_pred = TRUE)
)

view(xgb_results$.notes)

xgb_results

xgb_results %>% collect_metrics()

## VISUALIZING ACCURACY METRICS
xgb_results %>%
  collect_metrics() %>%
  filter(.metric == "roc_auc") %>%
  select(mean, mtry:sample_size) %>%
  pivot_longer(mtry:sample_size,
               values_to = "value",
               names_to = "parameter"
  ) %>%
  ggplot(aes(value, mean, color = parameter)) +
  geom_point(alpha = 0.8, show.legend = FALSE) +
  facet_wrap(~parameter, scales = "free_x") +
  labs(x = NULL, y = "AUC")

## BEST PERFORMING SET OF PARAMETERS
show_best(xgb_results, "accuracy")

## SELECT BEST ACCURACY MEASUREMENT
best_accuracy <- select_best(xgb_results, "accuracy")
best_accuracy

## FINALIZE WORKFLOW
xgb_final <- finalize_workflow(
  xgb_workflow,
  best_accuracy
)

## VIP PLOT
xgb_final %>%
  fit(data = soccer_train) %>%
  pull_workflow_fit() %>%
  vip(geom = "point")

## FIT TO TRAINING DATA
final_results <- last_fit(xgb_final, soccer_split)
collect_metrics(final_results)

#### STUFF TO SAVE
saveRDS(xgb_results, "./xgb_results.rds")
saveRDS(xgb_final, "./xgb_final.rds")
```


